---
title: "Association rule mining"
author: "Ali Khan"
date: "2023-08-14"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
library(igraph)
library(arules)
library(arulesViz)
library(knitr)

# Load the data
groceries_df <- read.delim("groceries.txt", header = FALSE, sep = ",", na.strings = c("", NaN))
groceries <- read.transactions("groceries.txt", format = "basket", sep = ",")
```

# Exploring the data set

The dataset has a total of 15296 items with a maximum of 4 items in the cart. 
Below is a snippet of the first few rows of the dataset.

```{r echo = FALSE}
kable(head(groceries_df), format = "markdown")
```

Next, we are going to observe the top 20 items that have been found in the 
baskets of each of the customs.

```{r echo = FALSE}
# Generate item frequency table
item_freq <- itemFrequency(groceries, type = "absolute")

# Convert item frequency to a data frame
item_freq_df <- data.frame(item = names(item_freq), frequency = item_freq)

# Sort by frequency in descending order and keep top N items
top_n_items <- 20
top_items <- head(item_freq_df[order(-item_freq_df$frequency), ], top_n_items)

# Create a horizontal bar plot using ggplot2
ggplot(top_items, aes(x = frequency, y = reorder(item, frequency))) +
  geom_bar(stat = "identity", fill = "#1f77b4") +
  ggtitle("Top Purchased Items in Groceries") +
  xlab("Occurrences") +
  ylab("Purchased Item") +
  theme_minimal() +
  theme(axis.text.y = element_text(hjust = 0, size = 10))
```

# Applying the Association Rule


## Initial Rule: 
Since this is the preliminary analysis, we are establishing the following rules:

  - the minimum fraction of carts that contain all of the items referenced in the 
rules is 0.1% (i.e support = 0.001)
  - the minimum confidence that we will find the item in the cart is 10% (conf = 0.1).
We are starting with a very low value just to get an idea of what the plot would
look like.
  - We want to take a look at the combination of at least two items hence, the 
minlen is 2.
```{r include = FALSE}
rules <- apriori(groceries,parameter=list(supp=0.001,conf=0.1,minlen=2))
```

Here's a table demonstrating the results of the rules that we have established.
```{r echo = FALSE}
inspect(rules[1:5])
```

```{r, echo=FALSE}
plot(rules, method = "scatter", measure = c("support", "confidence"), shading = "lift", jitter = 0)
```
We found that there were 32783 Rules. That is too high of a number! Let's prune 
it down and take a look at fewer rules.

## Adding Some Restrictions:
```{r, echo=FALSE}
sub_rules = subset(rules, subset=lift > 1 & confidence > 0.2)
plot(subset(rules, subset=lift > 1 & confidence > 0.2), measure = c("support", "confidence"), shading = "lift",jitter = 0)
```

Even after adding limitations where we are looking at lift>1 and confidence>0.2, 
we still have too many rules. Let's add some more restrictions and prune it 
further. 

## Final Analysis:
The new rules are as follows:


  - the minimum confidence that we will find the item in the cart is 70% (conf = 0.7).
If the confidence is very low (i.e 10%), we are saying that the possibility of 
someone buying a bread given that they are buying whole milk is only 10%. That 
wouldn't help us if we were for example reorganizing the store layout and 
putting items next to each other based on the likelihood of the customer buying 
the combination of the items.

  - The lift will be greater than 8. Meaning we are looking at the factor by 
which the probability has increased of finding an item on the LHS given that 
we know the item on the RHS.


```{r, echo=FALSE}
better_rules = subset(rules, subset=lift > 7 & confidence > 0.7)

plot(better_rules, measure = c("support", "confidence"), shading = "lift", jitter = 0)
```
```{r, echo=FALSE}
plot(better_rules, method='graph')
#plot(better_rules, method='graph', engine='htmlwidget')
```

Taking a look at the rule with the highest confidence (dark red circle), we can see that if we find liquor and red/blush wine, in the basket, we are likely to find bottled beer (there's a confidence of approximately 90.5%). In the same manner, we can read the rest of the rules where the items at the tail of the arrow (pointing towards the rule) represent items in the LHS while the items at the head of the arrow (pointing out of the rule) represent items on the RHS.
